---
title: "Responder_Fallacy"
author: "Anupam kumar Singh, MD"
date: "5 December 2017"
output: html_document
---

Stephen senn has written an excellent article about hype of precision medicine and how the responder-fallacy has led to false dawn of precisionomics. 
In a highly instructive article - he shows how use of term "response" is often abused and can mislead us. We will try to simulate this example in R

Let us consider a counter-factual experiment in which 1000 patients with mean FEV1 ( a measure of pulmonary function ) of 2.3 L and standard deviation of 0.26 L  are given an inhaler which increases each value on average by 0.2 L with  sd of 0.03L

```{r}
pre = rnorm(1000,2.3,0.26)
post= pre+ rnorm(1000,0.5,0.03)
```

Let's look at histogram of FEV1 at baseline
```{r}
library(tidyverse)
library(gridExtra)

inhale = data.frame(pre,post)

inhale %>% ggplot(aes(pre))+geom_histogram(fill="blue")+ geom_vline(xintercept = mean(pre))

```

Let's look at histogram of FEV1 in post(2nd reading)

```{r}
inhale %>% ggplot(aes(post))+geom_histogram(fill="blue")+ geom_vline(xintercept = mean(pre))

```

Now let's sample 24 patients from this population

```{r}
data_frame(placebo= pre[1:24], active= post[1:24]) %>% 
  mutate(id=row_number()) %>%  # add row_number as id
  gather(group,FEV1,-id) %>%  # note - sign to keep out id, key value neednt be explicit
  
  ggplot( aes(x=as.factor(id), y=FEV1)) + geom_point(aes(color=group)) + geom_line()+ labs( title= "Counter Factual trial",
                                                                                                             x = "Patient",
                                                                                                             y = "FEV1")+ coord_flip()
```

Here we can see individual response.
Now of course such a counterfactual trial is not possible , we have parallel group trial where each member belongs to either placebo or active treatment group

```{r,warning=FALSE}

data_frame(placebo= pre[1:24], active= post[1:24]) %>% 
  mutate(id=row_number()) %>%  # add row_number as id
  gather(group,FEV1,-id) %>%  # note - sign to keep out id, key value neednt be explicit
  
  filter((group=="placebo"&id%%2==0)|(group=="active"&id%%2!=0)) %>% 
  ggplot( aes(x=as.factor(id), y=FEV1)) + geom_point(aes(color=group)) +  labs( title= "parallel trial",
                                                                                                             x = "Patient",
                                                                                                             y = "FEV1")+ coord_flip()

```


Thus in this real world parallel trial it is very difficult to determine response as was seen in counterfactual trial 


So How do we determine response by crossover trials ..

Let's consider a crossover  trial with first crossover with active-placebo difference of 0.5 with sd of 0.2 L .

Now suppose we arbitrarily classify responder as those having more than 0.3 L increase


```{r}
set.seed(7)
crossover1 = rnorm(1000,0.5,0.2)

kil = ecdf(crossover1)(0.3)*100
data_frame(crossover1 = rnorm(1000,0.5,0.2)) %>% 
  ggplot(aes(crossover1))+geom_histogram(fill="blue")+geom_vline(xintercept = 0.3)+
  annotate("text",x=1,y=90,label=paste(kil," % non-responders"))


```

Thus we see that while every patient has some increase almost 14.8% will be labeelel non-responder due to arbitrary dichotomization , one of the reasons to avoid dichomotization which can lead to erroneous conclusions.


So how do we judge response then let's see the second  crossover trial in which there is an element of response i.e. the placebo-active difference is highly correlated(r=0.9) with first crossover trial 

```{r}
simcor = function(x, r,m,s){ r2 = r**2 
ve = 1-r2 
SD = sqrt(ve) 
e = rnorm(length(x), mean=0, sd=SD) 
y = r*x + e  
y1 = m+y*s

return(y1) } 


x=scale(crossover1)


crossover2 = simcor(x=x,r=0.9,m=0.5,s=0.2)

kil1 = ecdf(crossover2)(0.3)*100

data_frame(crossover2 = as.vector(crossover2)) %>% 
  ggplot(aes(crossover2))+geom_histogram(fill="blue")+geom_vline(xintercept = 0.3)+
  annotate("text",x=1,y=90,label=paste(kil1," % non-responders"))


```


So now we see 16.2% in second crossover trial( which are highly correlated with first cross-over trial) are non-responders( response being defined as greater than 0.3 L increase)


```{r}
p = data_frame(crossover1 = as.vector(crossover1),crossover2 = as.vector(crossover2)) %>% 
  mutate(group = case_when(
  crossover1 >0.3 & crossover2 >0.3 ~ "responders",
   crossover1 >0.3 & crossover2 <0.3 ~ "inconsistent",
  crossover1 <0.3 & crossover2 >0.3 ~"inconsistent",
  crossover1 <0.3 & crossover2 <0.3 ~ "non-responders"
  
)) %>% ggplot(aes(crossover1,crossover2,color=group))+geom_point()+geom_hline(yintercept=0.3)+geom_vline(xintercept=0.3)+geom_hline(yintercept=0.5)+geom_vline(xintercept=0.5)

p

ggExtra::ggMarginal(p, type = "histogram")


  

```



```{r}

data_frame(crossover1 = as.vector(crossover1),crossover2 = as.vector(crossover2)) %>% 
  mutate( response1 = case_when(crossover1<0.3 ~ "non-responder",
                                TRUE~"responder"),response2 = case_when(crossover2<0.3 ~ "non-responder",
                                TRUE~"responder")) %>% select(response1,response2) %>% group_by(response1,response2) %>% summarise(n=n(),percent = n() / nrow(.) )
```





Now consider a situation with no element of personal response

```{r}
crossover2 = simcor(x=x,r=0.02,m=0.5,s=0.2)


data_frame(crossover1 = as.vector(crossover1),crossover2 = as.vector(crossover2)) %>% mutate( response1 = case_when(crossover1<0.3 ~ "non-responder",
                                TRUE~"responder"),response2 = case_when(crossover2<0.3 ~ "non-responder",
                                TRUE~"responder")) %>% select(response1,response2) %>% group_by(response1,response2) %>% summarise(n=n(),percent = n() / nrow(.) )
```

```{r}
q = data_frame(crossover1 = as.vector(crossover1),crossover2 = as.vector(crossover2)) %>% mutate(group = case_when(
  crossover1 >0.3 & crossover2 >0.3 ~ "responders",
   crossover1 >0.3 & crossover2 <0.3 ~ "inconsistent",
  crossover1 <0.3 & crossover2 >0.3 ~"inconsistent",
  crossover1 <0.3 & crossover2 <0.3 ~ "non-responders"
  
)) %>% ggplot(aes(crossover1,crossover2,color=group))+geom_point()+geom_hline(yintercept=0.3)+geom_vline(xintercept=0.3)+geom_hline(yintercept=0.5)+geom_vline(xintercept=0.5)

q

ggExtra::ggMarginal(q, type = "histogram")


```

