---
title: "Regression to mean"
author: "Anupam kumar Singh, MD"
date: "7 November 2017"
output: html_document
---


This demonstration is based on regression to mean example by stephen senn in a [must read](https://www.google.co.in/url?sa=t&rct=j&q=&esrc=s&source=web&cd=1&cad=rja&uact=8&ved=0ahUKEwjC55u09vLXAhWJtY8KHRrfAtAQFggoMAA&url=http%3A%2F%2Fwww.dcscience.net%2FSenn-2009-Three-things-medical-writers.pdf&usg=AOvVaw3_pin00nZyrwodEaQvuNpq) paper by him.

Let us randomly sample baseline BP of 1000 patients with mean diastolic BP 90 mm Hg and standard deviation 11 mm Hg.

```{r}
set.seed(7)
baseline = rnorm(1000,90,11)

```

Now let's generate an outcome BP at second time point which is correlated (r=0.8) with first baseline measurement and has same mean diastolic BP and standard deviation  . you can also assume it is the second measurement after giving placebo.
```{r}
simcor = function(x, r,m,s){ r2 = r**2 
ve = 1-r2 
SD = sqrt(ve) 
e = rnorm(length(x), mean=0, sd=SD) 
y = r*x + e  
y1 = m+y*s

return(y1) } 


x=scale(baseline)

outcome=simcor(x=x,r=0.8,m=90,s=11)
```


Just to cross check, let's see mean , sd and correlation od outcome(post) with baseline(pre).
```{r}
sd(outcome)
mean(outcome)
cor(baseline,outcome)

```


Let's run a paired t-test between pre and post which will be non-significant as expected since mean DBPand sd are same and measurements are correlated
```{r}
t.test(baseline,outcome,paired=TRUE)
```


Let's generate a dataframe of these baseline and outcome variable and generate two more variables 
1. delta : difference in baseline and outcome
2. group : 1000 patients divided in three groups
a) hypertensives : Both baseline and outcome with DBP>95(arbitrary cut off)
b) inconsistent : Either baseline our outcome less than 95 on one occasion and higher than 95 on other
c) normotensives : Both baseline and outcome less than 95 on both occasions.

```{r,warning=FALSE}
df = data.frame(baseline,outcome)

library(dplyr)
library(ggplot2)

df = df %>% mutate(group = case_when(
  baseline<95 & outcome <95 ~ "normotensives",
   baseline>95 & outcome <95 ~ "inconsistent",
  baseline<95 & outcome >95 ~ "inconsistent",
  baseline>95 & outcome >95 ~ "hypertensives"
  
), delta=outcome-baseline)

```

Let's see a histogram of difference pre-post(baeline and outcome)

```{r,warning=FALSE}
ggplot(df,aes(x=delta))+geom_histogram(fill="blue")+geom_vline(xintercept=0)+geom_density()

```

Let's see a kernel density estimate

```{r}
ggplot(df,aes(x=delta))+geom_density()
```

Let's see frequency polygon based on groups

```{r}
ggplot(df,aes(x=delta))+geom_freqpoly(aes(color=group))
```

Let's see summary stats of difference

```{r}
df %>% select(delta) %>% summary()
```




Let's see a groupwise summary based on cut-off of 95

```{r}
library(tidyverse)
df %>% group_by(group) %>% summarize(n=n(),meanbaseline=mean(baseline),sdbaseline=sd(baseline),meanoutcome=mean(outcome),sdoutcome=sd(outcome),meandelta=mean(delta),sddelta=sd(delta)) %>%rownames_to_column %>% 
   gather(var, value, -rowname) %>% 
   spread(rowname, value) 
```



Lets visualise a plot of baseline vs outcome

```{r}
df %>% ggplot(aes(baseline,outcome,color=group))+geom_point()+geom_hline(yintercept=95)+geom_vline(xintercept=95)

```


Let's look at boxplot of pre(baseline) and post(outcome)

```{r}
library(tidyr)
df %>% select(baseline,outcome) %>% gather(key="group",value="DBP") %>% ggplot(aes(y=DBP,x=group))+geom_boxplot()
```

Now let's select only patients with DBP>95 this would be similar to hypertensive only patients entering an arm of control trial
```{r}
df %>% filter(baseline>95) %>% ggplot(aes(baseline,outcome,color=group))+geom_point() +geom_hline(yintercept=95)+geom_vline(xintercept=95)

```

```{r}
df %>% filter(baseline>95) %>% select(delta) %>% summary()
```
Thus we see in this selected group, outcome is lower than baseline as compared to previous group due to selection

Let's see histogram and density plots

```{r}
df %>% filter(baseline>95) %>% ggplot(aes(x=delta))+geom_histogram(fill="blue")+geom_vline(xintercept=0)+geom_density()
```



```{r}
df %>% filter(baseline>95) %>% ggplot(aes(x=delta))+geom_density()
```

We run the paired t test and see that it is significant now.

```{r}
df1 = df %>% filter(baseline>95)
t.test(df1$outcome,df1$baseline,paired = TRUE)
```

```{r}
df %>% select(baseline,outcome) %>% filter(baseline>95) %>% gather(key="group",value="DBP") %>% ggplot(aes(y=DBP,x=group))+geom_boxplot()
```


### KEY POINTS

1. Scientists should be very careful while interpreting effect of drug/change in parameter based on a single arm
2. Control group is important to address regression to mean.
3. Regression to mean accounts contributes a lot to so-called "placebo effect" or "alternative medicine".

## Reference
1. [Three things that every medical writer should know about statistics](https://www.google.co.in/url?sa=t&rct=j&q=&esrc=s&source=web&cd=2&cad=rja&uact=8&ved=0ahUKEwjC55u09vLXAhWJtY8KHRrfAtAQFggxMAE&url=http%3A%2F%2Feprints.gla.ac.uk%2F8107%2F1%2Fid8107.pdf&usg=AOvVaw1Y1KazmlhlVrw03fawlUrZ)


